Forcing for Isabelle

### 2019-01-05 LICS/proof-of-separation.tex
breve reporte sobre powerset

We now turn to the Powerset Axiom. We followed the proof of
\cite[IV.2.27]{kunen2011set}, and actually the main technical result,
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ Pow{\isacharunderscore}inter{\isacharunderscore}MG{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}a{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}Pow{\isacharparenleft}a{\isacharparenright}\ {\isasyminter}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
keeps most of the structure of the printed proof; this ``skeleton'' of the
argument takes around 120 (short) lines. There are approximately 30
more lines of bureaucracy in the proof of the last lemma. We preserved the
names of variables used in the textbook, with the occasional question
mark that distinguishes meta-variables.

Two more
absoluteness results concerning powersets were needed: These are
refinements of results located in \isatt{Relative} that assume that
$y$ below is in the class $M$.
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}subset{\isacharunderscore}Pow{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}y\ {\isasymsubseteq}\ Pow{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}abs{\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}\ {\isasymlongleftrightarrow}\ y\ {\isacharequal}\ {\isacharbraceleft}a{\isasymin}Pow{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ M{\isacharparenleft}a{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

>>>>

We now turn to the Powerset Axiom. We followed the proof of
\cite[IV.2.27]{kunen2011set}, to which we refer the reader for further
details, trying to  preserve the
names of variables used in the textbook (with the occasional question
mark that distinguishes meta-variables). Actually, the main technical result
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ Pow{\isacharunderscore}inter{\isacharunderscore}MG{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}a{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}Pow{\isacharparenleft}a{\isacharparenright}\ {\isasyminter}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
keeps most of the structure of the printed proof; this ``skeleton'' of the
argument takes around 120 (short) lines. There are approximately 30
more lines of bureaucracy in the proof of the last lemma.

Two more
absoluteness lemmas concerning powersets were needed: These are
refinements of results (\isatt{powerset\_Pow} and
\isatt{powerset\_imp\_subset\_Pow}) located in the theory \isatt{Relative} which assume that
$y$ below is in the class $M$.
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}subset{\isacharunderscore}Pow{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}y\ {\isasymsubseteq}\ Pow{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}abs{\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}\ {\isasymlongleftrightarrow}\ y\ {\isacharequal}\ {\isacharbraceleft}a{\isasymin}Pow{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ M{\isacharparenleft}a{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

Of the rest of the theory file \verb|Powerset_Axiom.thy|, a
considerable fraction is taken by the proof of a closure property of
the ctm $M$, that involves renaming of an internalized formula; also,
the handling of the projections \isatt{fst} and \isatt{snd} must be
done internally. 
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ sats{\isacharunderscore}fst{\isacharunderscore}snd{\isacharunderscore}in{\isacharunderscore}M{\isacharcolon}\isanewline
\ \isakeyword{assumes}\ \isanewline
\  \ {\isachardoublequoteopen}A{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}B{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}p{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}l{\isasymin}M{\isachardoublequoteclose}\isanewline
\ \  {\isachardoublequoteopen}o{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymchi}{\isasymin}M{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}\ {\isasymle}\ {\isadigit{6}}{\isachardoublequoteclose}\isanewline
\  \isakeyword{shows}\isanewline
\  \ {\isachardoublequoteopen}{\isacharbraceleft}sq\ {\isasymin}\ A{\isasymtimes}B\ {\isachardot}\isanewline
\ \ \ \ \ sats{\isacharparenleft}M{\isacharcomma}{\isasymphi}{\isacharcomma}{\isacharbrackleft}p{\isacharcomma}l{\isacharcomma}o{\isacharcomma}snd{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}fst{\isacharparenleft}sq{\isacharparenright}{\isacharcomma}{\isasymchi}{\isacharbrackright}{\isacharparenright}{\isacharbraceright}\ {\isasymin}\ M{\isachardoublequoteclose}\ \isanewline
\end{isabelle}
